Framing Effects in the Wild

### 2022-08-25 paper/Manuscript_Framing.Rmd
Incorporate TM feedback (2/2)

Research on framing has shown for decades that citizens' attitudes and their issue associations are highly reactive to minor changes in the emphasis of specific considerations [@Chong2007; @Chong2010; @Druckman2012Bias; @Kinder1990; @Nelson1996; @Nelson1997; @sniderman2004structure; for reviews see @Busby2019; @Druckman2001; @Leeper2020]. However, it remains unclear how well these experimental findings translate into the real world. First, taking experimental effects as estimates of the real-world impact of framing means making strong assumptions about citizens' exposure to frames: while framing studies usually simply present a news article or other forms of political content to respondents, news content in the real world competes with other news and entertainment content to reach its audience. Changes in news framing itself might affect how likely readers are to consume news content: readers might be less likely to consume frames challenging their existing attitudes and reject news outlets containing such frames (and vice versa seek out attitudinally congruent information). This idea of 'selective exposure/avoidance' [@Bennett2008], or a 'confirmation bias' [@Taber2006] brings with it a major endogeneity problem making purely correlational evidence of framing effects rather uninformative.

@@

To provide a baseline expectation regarding the effects, it is useful to consider the expected effect size at this point. Despite the abundance of framing effects studies, there is no common standard to assess the magnitude of framing effects. Even the correct measurement of the magnitude is a contested question, as different control groups (untreated, opposing frame, ...) can be compared [@Druckman2001; @Chong2007a, 109f for a discussion]. Additionally, framing effects differ based on the polarization and salience of the issue [@Lecheler2009a], repeated exposure [@Chong2007competitiveFraming], the sender [@Slothuus2010], political knowledge [@Chong2007b], as well as the reference of certain groups [@Nelson1996]. Generally, reported effect size around 30% of the scale are not uncommon [@Slothuus2010], sometimes even over 40% [@Chong2007a, 104], but most results lie in between 10% and 25% of the scale when comparing pro and con arguments\footnote{As many studies do not include a control group, this is the most comparable metric.} [@Lecheler2009a; @Nelson1997Psychology; @Nelson1997Rally; @Slothuus2008; @Slothuus2010]. As experimental treatments are likely designed in order to elicit strong responses, and given biases towards null effects in the production of scientific knowledge [see e.g. the recent debate regarding nudging effects: @Maier2022], reported effects likely display a selection bias towards strong treatments. The treatment can be considered rather strong here, as the crime consideration is associated with strongly negative evaluations and the relative increase in the prevalence of this content is large \footnote{6 percentage point increase compared to pre-treatment share of 8.36% in Bild, an over 70% increase.}. Additionally, given that readers already consumed the newspaper, the qualities of the sender should rather increase the effect size, while the high salience of the issue should decrease it. Given these considerations, I expect an effect of around 10% of the scale, or 0.6 on the seven-point scale\footnote{As the scale runs from one to seven, it spans six points wide.}, which represents a conservative estimate compared to reported experimental effects.

@@

```{r singleplot, fig.cap="Effect of change in framing on migration attitudes; difference-in-differences estimate. Dashed vertical lines indicate conservative experimental effect (0.6, baseline expectation), dotted vertical lines indicate minimal relevant effect (0.15). Full model visible in Appendix XX", fig.height=2}

@@

As outlined above, framing theory suggests that increased exposure to a specific consideration about an issue leads to stronger association of the issue with this consideration. In the present case, this should mean that the correlation of attitudes on immigration with attitudes about crime should increase. The GLES Panel data contains a variable asking respondents whether "the state's capabilities to fight crime should be extended, even if that would entail more surveillance", asked in three waves overlapping with the migration attitude question. Figure \@ref(fig:corplot) shows the correlation of crime- and migration attitudes in treatment and control group, pre- and post-treatment. I added a counterfactual, applying the change in the control group to the treatment group, for easier interpretation. As we can see, the treatment group actually saw a *smaller* increase in the correlation of crime and migration attitudes than the control group, *despite weaker pre-treatment association*. Following [@Nelson1996], table \@ref(tab:cortable) regresses immigration attitudes on attitudes towards crime and investigates whether this association is generally different between the two groups\footnote{These differences are however in the opposite direction as the (substantial) differences in the descriptive correlations outlined in the plot. I am currently not sure how to interpret this.}, changes post-treatment, and whether it changes more in the treatment group using interaction terms. I find minor differences between the groups, but no significant difference when interacted with the DiD-term - respondents do not seem to associate crime more strongly with migration following the change in migration coverage in *Bild*.

@@

@@

One reason for this null effect could be that a slowly unfolding long term effect is masked by initial absence of change. To assess whether an effect develops over time, figure \@ref(fig:migwaves) assesses the effect within split-samples for each post-wave compared to the pre-treatment wave. This allows the assessment of heterogeneous treatment effects across time. The black dots with associated confidence intervals indicate the estimated effect for each wave. The horizontal dashed line indicates the expected effect based on experimental evidence (0.6), whereas the dotted horizontal line indicates the SESOI of 0.15. The vertical (red) dashed line indicates the timing of the editorial change, whereas the dotted (orange) vertical line indicates the timing of the strong increase in crime framing in June 2017, several months after the takeover by Reichelt (see figure \@ref(fig:treatment)). It s visible that, over the course of the entire year 2017, the effect is clearly smaller than the SESOI, and includes zero. However, estimates for waves 7 and 8 in late 2018 and mid-2019 - more than one and a half years after the editorial change - showcase effects statistically different from zero, and including the SESOI boundary. Note however that the effect is still tiny, with a point estimate smaller than a 0.1 on the seven-point Likert scale, as well as an upper boundary just above the SESOI at ??. The waves in late 2019 and early 2020 have point estimates at similar lower levels (??, ??), but include zero again. Overall, this suggests an effect that remains below an effect size of minimal relevance.

```{r migwaves, fig.cap="Effect of Change in Framing on Immigration Attitudes across Waves"}

@@

While immigration attitudes remain unaffected, it might well be that an increased exposure to crime framing increases individual perceptions of the importance of migration, despite not changing attitudes, or that it raises concerns about the integration rather than the immigration of migrants. Similarly, support for the radical-right *AfD* as owner of the migration issue might increased as a result of increased exposure to crime frames. Figure \@ref(fig:multidv) displays standardized effects of the change in newspaper framing on immigration attitudes (as reference) and three related dependent variables, namely integration attitudes, the likelihood to report migration as the most important problem facing the country, as well as thermometer scores for the radical-right AfD (see Appendix ?? for precise variable descriptions). Dotted lines indicate a 0.2 standard deviation change, which has been defined as the standardized equivalence boundary in existing research [@Peyton2020, 599]. All estimated effects are smaller in size than 0.05 standard deviations and associated 95%-confidence intervals contain zero. This means we can reject the hypothesis of meaningful effect size, but cannot reject that the estimate is statistically different from zero. All estimates suggest a clear null effect.

```{r multidv, fig.cap="Standardised Effect of Change in Framing on Alternative Dependent Variables"}

@@

The presented results are robust evidence that, despite an over 70% increase of the emphasis of crime in migration content of Germany's largest newspaper, its readers migration attitudes remained surprisingly unaffected. Effect estimates remain below minimal thresholds across time. Related dependent variables such as integration attitudes, reporting migration as the most important problem, and support for the radical-right AfD do not deviate from those of other readers. Migration attitudes are not increasingly associated with crime attitudes in this group, nor are liberal Bild readers more likely to stop reading the newspaper, nor conservative non-readers more likely to report reading it.

@@

While these findings are worrisome for framing scholars, they are generally good news for democracy: even when outlets try, they cannot simply manipulate their audience's political attitudes with a snap of their finger. Worries about the undermining of democracy trough the takeover of partisan news outlets might be overstated, as consumers either seem to hold a critical distance to their news outlets or consume content from so many outlets that a single one has not much influence. Descriptive data about American news diets supports this interpretation [@Guess2021a].

This study suffers from several limitations regarding the generalizabilty of the case at hand. First, the immigration issue was already highly salient for one and a half years when Reichelt took over *Bild*. This means that attitudes have possibly crystallized as the result of personal reasoning, exposure to information, or cues from one's preferred party. Second, while tabloids are often seen as manipulators of public opinion [@Foos2022], they also contain less political content and might be taken less seriously when they write about politics. Readers from other outlets might react differently. Lastly, Bild readers differ from other readers in many ways, among others they are more conservative and less knowledgeable about politics (see Appendix ??). Readers of other outlets might react differently to increased exposure to crime frames.

@@

>>>>

Research on framing has shown for decades that citizens' attitudes and their issue associations are highly reactive to minor changes in the emphasis of specific considerations [@Chong2007b; @Chong2010; @Druckman2012Bias; @Kinder1990; @Nelson1996; @Nelson1997Rally; @sniderman2004structure; for reviews see @Busby2019; @Druckman2001; @Leeper2020]. However, it remains unclear how well these experimental findings translate into the real world. First, taking experimental effects as estimates of the real-world impact of framing means making strong assumptions about citizens' exposure to frames: while framing studies usually simply present a news article or other forms of political content to respondents, news content in the real world competes with other news and entertainment content to reach its audience. Changes in news framing itself might affect how likely readers are to consume news content: readers might be less likely to consume frames challenging their existing attitudes and reject news outlets containing such frames (and vice versa seek out attitudinally congruent information). This idea of 'selective exposure/avoidance' [@Bennett2008], or a 'confirmation bias' [@Taber2006] brings with it a major endogeneity problem making purely correlational evidence of framing effects rather uninformative.

@@

To provide a baseline expectation regarding the effects, it is useful to consider the expected effect size at this point. Despite the abundance of framing effects studies, there is no common standard to assess the magnitude of framing effects. Even the correct measurement of the magnitude is a contested question, as different control groups (untreated, opposing frame, ...) can be compared [@Druckman2001; @Chong2007a, 109f for a discussion]. Additionally, framing effects differ based on the polarization and salience of the issue [@Lecheler2009b], repeated exposure [@Chong2007competitiveFraming], the sender [@Slothuus2010], political knowledge [@Chong2007b], as well as the reference of certain groups [@Nelson1996]. Generally, reported effect size around 30% of the scale are not uncommon [@Slothuus2010], sometimes even over 40% [@Chong2007a, 104], but most results lie in between 10% and 25% of the scale when comparing pro and con arguments\footnote{As many studies do not include a control group, this is the most comparable metric.} [@Lecheler2009b; @Nelson1997Psychology; @Nelson1997Rally; @Slothuus2008; @Slothuus2010]. As experimental treatments are likely designed in order to elicit strong responses, and given biases towards null effects in the production of scientific knowledge [see e.g. the recent debate regarding nudging effects: @Maier2022], reported effects likely display a selection bias towards strong treatments. The treatment can be considered rather strong here, as the crime consideration is associated with strongly negative evaluations and the relative increase in the prevalence of this content is large \footnote{6 percentage point increase compared to pre-treatment share of 8.36% in Bild, an over 70% increase.}. Additionally, given that readers already consumed the newspaper, the qualities of the sender should rather increase the effect size, while the high salience of the issue should decrease it. Given these considerations, I expect an effect of around 10% of the scale, or 0.6 on the seven-point scale\footnote{As the scale runs from one to seven, it spans six points wide.}, which represents a conservative estimate compared to reported experimental effects.

@@

```{r singleplot, fig.cap="Effect of change in framing on migration attitudes; difference-in-differences estimate. Dashed vertical lines indicate conservative experimental effect (0.6, baseline expectation), dotted vertical lines indicate minimal relevant effect (0.15). Full model visible in Appendix XX.", fig.height=2}

@@

As outlined above, framing theory suggests that increased exposure to a specific consideration about an issue leads to stronger association of the issue with this consideration. In the present case, this should mean that the correlation of attitudes on immigration with attitudes about crime should increase. The GLES Panel data contains a variable asking respondents whether "the state's capabilities to fight crime should be extended, even if that would entail more surveillance", asked in three waves overlapping with the migration attitude question.

Figure \@ref(fig:corplot) shows the correlation of crime- and migration attitudes in treatment (Bild-readers in pre-wave) and control group (never Bild-readers), pre- and post-treatment. I added a counterfactual, applying the change in the control group to the treatment group, for easier interpretation. As we can see, the treatment group actually saw a *smaller* increase in the correlation of crime and migration attitudes than the control group, *despite weaker pre-treatment association*. Following [@Nelson1996], table \@ref(tab:cortable) regresses immigration attitudes on attitudes towards crime and investigates whether this association is generally different between the two groups\footnote{These differences are however in the opposite direction as the (substantial) differences in the descriptive correlations outlined in the plot. I am currently not sure how to interpret this.}, changes post-treatment, and whether it changes more in the treatment group using interaction terms. I find minor differences between the groups, but no significant difference when interacted with the DiD-term - respondents do not seem to associate crime more strongly with migration following the change in migration coverage in *Bild*.

@@

[explain why this is done, and why constituent terms are not interesting - only the change in association of crime attitude and dependent variable is of interest, we are not trying to explain variation in the dependent variable! (which is also why R2 doesn't matter here)]

@@

If readers only perceive the change in content across longer time periods, the effect should slowly unfold across time. If that is the case, initial absence of change might bias the estimates downward and thereby mask an effect. To assess whether an effect develops over time, figure \@ref(fig:migwaves) assesses the effect within split-samples for each post-wave compared to the pre-treatment wave. This allows the assessment of heterogeneous treatment effects across time. The black dots with associated confidence intervals indicate the estimated effect for each wave. The horizontal dashed line indicates the expected effect based on experimental evidence (0.6), whereas the dotted horizontal line indicates the SESOI of 0.15. The vertical (red) dashed line indicates the timing of the editorial change, whereas the dotted (orange) vertical line indicates the timing of the strong increase in crime framing in June 2017, several months after the takeover by Reichelt (see figure \@ref(fig:treatment)). Over the course of the entire year 2017, the effect is clearly smaller than the SESOI, and includes zero. However, estimates for waves 7 and 8 in late 2018 and mid-2019 - more than one and a half years after the editorial change - showcase effects statistically different from zero, and including the SESOI boundary. Note however that the effect is still tiny, with a point estimate smaller than a 0.1 on the seven-point Likert scale, as well as an upper boundary just above the SESOI at ??. The waves in late 2019 and early 2020 have point estimates at similar lower levels (??, ??), but include zero again. Overall, this suggests an effect that remains below an effect size of minimal relevance.

```{r migwaves, fig.cap="Effect of Change in Framing on Immigration Attitudes across Waves. Dashed (black) horizontal lines indicate conservative experimental effect (0.6, baseline expectation), dotted horizontal lines indicate minimal relevant effect (0.15). Dashed (red) vertical line indicates effect of Reichelt takeover, dotted (orange) line timing of change in coverage. Full models visible in Appendix XX."}

@@

While immigration attitudes remain unaffected, it might well be that an increased exposure to crime framing affects other variables. For example, increased attention to crime within migration coverage might increase individual perceptions of the importance of migration, despite not changing attitudes, or raise concerns about the integration rather than the immigration of migrants. Similarly, support for the radical-right *AfD* as owner of the migration issue could increase as a result of increased exposure to crime frames. All variable descriptions and models can be found in the appendix.

Figure \@ref(fig:multidv) displays standardized effects of the change in newspaper framing on immigration attitudes (as reference) and three related dependent variables, namely integration attitudes, the likelihood to report migration as the most important problem facing the country, as well as thermometer scores for the radical-right AfD (see Appendix ?? for precise variable descriptions). Dotted lines indicate a 0.2 standard deviation change, which has been defined as the standardized equivalence boundary in existing research [@Peyton2020, 599]. All estimated effects are smaller in size than 0.05 standard deviations and associated 95%-confidence intervals contain zero. This means we can reject the hypothesis of meaningful effect size, but cannot reject that the estimate is statistically different from zero. All estimates suggest a clear null effect.

```{r multidv, fig.cap="Standardised effect of change in framing on alternative dependent variables. Dotted vertical lines indicate minimal relevant effect (0.2 SD). Full models available in Appendix XX."}

@@

The presented results are robust evidence that, despite an over 70% increase of the emphasis of crime in migration content of Germany's largest newspaper, its readers' migration attitudes remained surprisingly unaffected. Effect estimates remain below minimal thresholds across time. Related dependent variables such as integration attitudes, reporting migration as the most important problem, and support for the radical-right AfD do not deviate from those of other readers. Migration attitudes are not increasingly associated with crime attitudes in this group, nor are liberal Bild readers more likely to stop reading the newspaper, nor conservative non-readers more likely to report reading it.

@@

While these findings substantially question evidence from experimental framing studies, they are generally good news for democracy: even when outlets try, they cannot simply manipulate their audience's political attitudes with a snap of their finger. Worries about the undermining of democracy through the takeover of partisan news outlets might be overstated, as consumers either seem to hold a critical distance to their news outlets or consume content from so many outlets that a single one has not much influence. Descriptive data about American news diets supports this interpretation [@Guess2021a].

Despite the unique case and the extensive efforts to isolate the effect of the documented change inframing, this study suffers from several limitations regarding the generalizabilty of the case at hand. First, the immigration issue was already highly salient for one and a half years when Reichelt took over *Bild*. This means that attitudes have possibly crystallized as the result of personal reasoning, exposure to information, or cues from one's preferred party. Second, while tabloids are often seen as manipulators of public opinion [@Foos2022], they also contain less political content and might be taken less seriously when they write about politics. Readers from other outlets might react differently. Lastly, Bild readers differ from other readers in many ways, among others they are more conservative and less knowledgeable about politics (see Appendix ??). Readers of other outlets might react differently to increased exposure to crime frames.

[add: no randomisation of treatment!! plus adapt to new evidence regarding attitude stability]

@@

\pagebreak
